---
title: "The `anova_*` functions"
subtitle: "What they do, and the thinking behind them"
author: "Ryan Dobson"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: false
    theme: cosmo
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
    message: false
    warning: false
    fig-width: 8
    fig-height: 5
    dpi: 110
---

```{r}
#| label: setup
#| include: false
library(codechest)
library(ggplot2)

# One console width for the whole document. The printed output wraps to
# getOption("width"), so setting it here keeps every block sharing a right
# edge instead of each ending wherever R happened to default.
options(width = 76)
```

## The idea in three sentences

A complete, defensible one-way ANOVA takes a page of code: descriptives, two
or three assumption checks, the test, an effect size, group means with the
right standard error, a figure, and a sentence. Most of that page is the same
every time, and the parts that vary are decisions you should be making
deliberately rather than retyping.

These functions do the routine parts in one line each, and make the
decisions visible: **what you should do next depends on what the assumption
checks found**, so the checks are a separate step that ends in a
recommendation with a citation attached.

Everything below runs on `PlantGrowth`, which ships with R, so anything here
can be pasted and run.

## Assumptions first, and separately

This is the design decision everything else follows from.

Assumption checking is not a formality you complete before the "real"
analysis. It *determines* which analysis is the right one. So it is its own
function, and it ends by telling you what to run.

```{r}
#| label: check
fit <- aov(weight ~ group, data = PlantGrowth)
anova_check(fit)
```

Three things to notice.

**Normality is tested on the residuals, not the raw scores.** The raw values
pool three groups with different means, so they can look non-normal when every
group is perfectly normal. This is the single most common error in student
work, and it is not something you should have to remember.

**Independence is reported as "not testable" rather than passing.** It is a
property of how the study was run. No test can confirm it, so the function
does not pretend otherwise.

**The output ends with a recommendation, not a verdict.** Which brings us to
the point.

### The recommendation changes with the data

Same one line of code. Different data. Different advice.

```{r}
#| label: unequal-variances
# unequal variances AND unequal n: the case that actually distorts F
set.seed(11)
d <- data.frame(
  g = factor(rep(c("a", "b", "c"), times = c(30, 10, 10))),
  y = c(rnorm(30, 10, 1), rnorm(10, 11, 6), rnorm(10, 12, 6))
)
print(anova_check(y ~ g, data = d), verbosity = "short")
```

```{r}
#| label: non-normal
# skewed residuals in small groups
set.seed(13)
d2 <- data.frame(
  g = factor(rep(c("a", "b", "c"), each = 10)),
  y = c(rexp(10, .2), rexp(10, .2) + 1, rexp(10, .2) + 2)
)
print(anova_check(y ~ g, data = d2), verbosity = "short")
```

That second recommendation is deliberately *not* "transform the outcome."
Field and Wilcox (2017) give six reasons transformations are seldom worth it,
the sharpest being that transforming changes the hypothesis you are testing.
They are also pointed that trimming at 2.5 *SD* and then running OLS is worse
than doing nothing, because the *SD* is itself pulled by the outliers, so they
get masked, and the standard errors afterwards are wrong at any sample size.

The recommendation rules live in one table rather than being scattered through
the tests, so what the package advises can be read, argued with, and changed
in one place.

## Every claim carries its citation

Each recommendation names the concepts it rests on. Those are keys into a
reference registry, so the reasoning is always one call away.

```{r}
#| label: ref-homogeneity
anova_ref("homogeneity")
```

```{r}
#| label: ref-omega
anova_ref("omega_squared")
```

Equation numbers are from Maxwell, Delaney and Kelly (2017, 3rd ed.), read out
of the book rather than remembered. Where an equation number has not been
confirmed it is left blank rather than guessed, on the principle that a wrong
equation number is worse than a missing one, because a reader cannot tell it
is wrong.

```{r}
#| label: refs-todo
anova_refs_todo()
```

## The analysis

Once the checks say to proceed:

```{r}
#| label: report
anova_report(fit)
```

Two effect sizes, because the distinction matters and is easy to get wrong.
$R^2$ (what most of psychology calls $\eta^2$ in this context) describes the
sample and is biased upward as an estimate of the population value.
$\hat{\omega}^2$ corrects that bias and is the one to report.

$$\hat{\omega}^2 = \frac{SS_{effect} - df_{effect} \cdot MS_W}{SS_{total} + MS_W}$$

## Group means, with the right standard error

```{r}
#| label: means
anova_means(fit)
```

The standard error is the **model-based** one:

$$SE(\bar{Y_j}) = \sqrt{\frac{MS_W}{n_j}}$$

not each group's own $s_j/\sqrt{n_j}$. That is the standard error consistent
with the *F* test being reported alongside it, so error bars built from it
match the model. Because the design is balanced, all three come out equal,
which is the model saying it has one variance estimate and equal *n*.

`pooled = FALSE` gives the per-group version, which is what you want when
homogeneity is violated.

## The figure

```{r}
#| label: figure
anova_plot(fit, title = "Plant growth by treatment")
```

The message underneath is not decoration. Figures are where quiet decisions
get made, so the ones this function made are stated, along with the line of
code to change them.

The axis spans the data **and** the confidence intervals plus 10%, rather than
zooming tight to the data. A truncated axis exaggerates differences between
means. For an outcome with a real measurement scale, say a 1–7 Likert item,
pass it and it is used exactly, with every point labelled:

```{r}
#| label: likert
set.seed(5)
lik <- data.frame(
  cond   = factor(rep(c("control", "primed"), each = 40)),
  rating = c(sample(3:6, 40, TRUE), sample(4:7, 40, TRUE))
)
anova_plot(aov(rating ~ cond, data = lik), y_range = c(1, 7),
           title = "A 1-7 item, shown on its own scale", quiet = TRUE)
```

Limits are applied with `coord_cartesian()`, which zooms. The obvious
alternative, `scale_y_continuous(limits = )`, silently *drops* observations
outside the range, which would change the violin and the intervals rather than
just the view of them.

## Does the conclusion survive the decisions you could have made?

Field and Wilcox's central recommendation is that sensitivity analysis should
be routine: fit the conventional model and a robust counterpart, report either
where they agree, report the robust one where they do not. Their stronger
claim is that this comparison is the only known way to judge whether a
conventional method gave a reasonable answer.

`anova_multi()` automates that and widens it, crossing the small analysis
decisions you could defensibly have made.

```{r}
#| label: multiverse
set.seed(8)
sk <- data.frame(
  g = factor(rep(c("ctrl", "trt"), each = 20)),
  y = c(rlnorm(20, 2.0, .85), rlnorm(20, 2.45, .85))
)
mv <- anova_multi(y ~ g, data = sk, contrast = c("trt", "ctrl"))
mv
```

Here the conventional analysis gives *p* = .010 and the trimmed-means version
gives *p* = .051. That is exactly the situation the tool exists to surface: a
conclusion resting on one defensible decision.

```{r}
#| label: forest
#| fig-height: 5.5
anova_multi_plot(mv)
```

The `CI vs base` column is the price of robustness in the only currency that
matters here: how much wider the interval got.

## Repeated measures

Within-subjects designs need long-format data and a subject identifier. The
wide matrix needed for the sphericity test is built internally, so nothing is
reshaped by hand.

```{r}
#| label: within
set.seed(9)
n <- 20
subj <- rnorm(n, 0, 5)
long <- data.frame(
  participant = factor(rep(seq_len(n), 4)),
  block = factor(rep(paste0("b", 1:4), each = n)),
  score = c(subj + rnorm(n, 20, 2), subj + rnorm(n, 22, 9),
            subj + rnorm(n, 24, 2), subj + rnorm(n, 25, 14))
)
print(anova_check(score ~ block, data = long, id = "participant"),
      verbosity = "short")
```

Sphericity is checked, both epsilons are reported, and the correction is
chosen on the usual $\varepsilon < .75$ convention rather than left to you.

## How it is put together

Six layers, and the separation is the point:

| Layer | Does |
|---|---|
| Orchestration | `anova_check()`, `anova_workup()`: what you call |
| Design engines | one per design, knowing which assumptions apply |
| Tests | each returns a *finding*: a result plus a severity |
| Knowledge base | the reference registry and the recommendation rules |
| Presentation | print methods, figures, the APA sentence |
| Multiverse | `anova_multi()` |

The invariant that keeps it honest: **a test produces a finding; a finding
does not know what to do about itself.** Changing what the package advises
about unequal variances means editing one rule, not Levene's test.

Everything is base R inside, plus ggplot2 for figures. Levene's test is
implemented directly rather than importing `car`, and is verified against
`car::leveneTest()` to ten decimal places. The Greenhouse–Geisser and
Huynh–Feldt epsilons are computed from the covariance matrix and verified
against R's own `anova.mlm`.

## What is not built yet

Being explicit about this, because a tool that quietly does the wrong thing is
worse than one that refuses:

- **Factorial and mixed designs.** Not implemented. `anova_check()` will tell
  you so rather than guess.
- **Generalized $\eta^2$.** The right effect size for repeated-measures and
  factorial designs. Until it exists, `anova_report()` **refuses** those
  models rather than returning a plausible-looking wrong number.
- **Post hoc comparisons.** Not yet. The omnibus test does not license claims
  about specific pairs, and the package does not pretend otherwise.

## References

Field, A. P., & Wilcox, R. R. (2017). Robust statistical methods: A primer for
clinical psychology and experimental psychopathology researchers. *Behaviour
Research and Therapy, 98*, 19–38.

Lakens, D. (2013). Calculating and reporting effect sizes to facilitate
cumulative science: A practical primer for t-tests and ANOVAs. *Frontiers in
Psychology, 4*, 863.

Maxwell, S. E., Delaney, H. D., & Kelly, K. (2017). *Designing experiments and
analyzing data: A model comparison perspective* (3rd ed.). Taylor & Francis.

```{r}
#| label: textbook
#| echo: false
cat(strwrap(anova_textbook(), width = 73), sep = "
")
```

Every other reference the package relies on is in the registry; `anova_ref()`
with no argument lists them all.
